---
title: 'Class 10: Halloween Candy'
author: "Angelita Rivera (PID A15522236)"
date: "10/28/2021"
output: html_document
---

#Get the data! 

Here we look at 538 Halloween candy data. We take the CSV file from their GitHub repo. 

```{r}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy <- read.csv(url, row.names = 1)
head(candy, n =5)
```

> **Q1.** How many different candy types are in this dataset?

```{r}
dim(candy)
```
There are 85 different pieces of candy in this dataset. 

> **Q2.** How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types in this data set. 

> **Q3.** What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy
```
Peanut M&Ms are my favorite candy in the dataset. 

```{r}
candy["Peanut M&Ms",]$winpercent
```
It's winpercent value is 69.48379. 

> **Q4.** What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

> **Q5.** What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

> **Q6.** Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
library("skimr")
skim(candy)
```

Winpercent is on a different scale than the rest of the columns. 

> **Q7.** What do you think a zero and one represent for the candy$chocolate column?

Since the chocolate column says whether or not it has chocolate; I would think that a "1" or "TRUE" means it does, while a "0" or "FALSE" means it doesn't. 

> **Q8.** Plot a histogram of winpercent values

```{r}
library(ggplot2)
hist(candy$winpercent)
```

> **Q9.** Is the distribution of winpercent values symmetrical?

No, the histogram is not symmetrical. The plot is skewed to the right. 

> **Q10.** Is the center of the distribution above or below 50%?

It is below 50%. 

> **Q11.** On average is chocolate candy higher or lower ranked than fruit candy?

First, we need to find all the chocolate candy rows in the 'candy' data set. 
```{r}
#Turn into TRUE's and FALSE's
inds <- as.logical(candy$chocolate)
inds.chocolate <- candy[inds,]$winpercent
mean(inds.chocolate)
```

Now, let's do it for fruity candy: 
```{r}
fruity <- as.logical(candy$fruity)
inds.fruity <- candy[fruity,]$winpercent
mean(inds.fruity)
```

Chocolate candy is ranked higher than fruity candy; as it has a mean value of 60.92153 compared to 44.11974.

> **Q12.** Is this difference statistically significant?

```{r}
t.test(inds.chocolate, inds.fruity)
```

The difference is significant, as we have a p-value of 2.871e-08. Chocolate is preferred over fruity candy.

> **Q13.** What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```
The five least liked candy types in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble and Jawbusters. 

> **Q14.** What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(-(candy$winpercent)),], n=5)
```
The top 5 all time favorite candy types out of this set are ReeseOs Peanut Butter cups, ReesesOs Miniatures, Twix, Kit Kat and Snickers. 

> **Q15.** Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> **Q16** This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> **Q17.** What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets. 

> **Q18.** What is the best ranked fruity candy?

The best ranked fruity candy is Starbursts.

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols)
```

> **Q19.** Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

According to the plot, it appears ReesesOs Miniatures has the highest winpercent and the lowest pricepoint. 

> **Q20.** What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> **Q22.** Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Choocolate and fruity have the highest anti-correlation; but bar and pluribus, and bar and fruity also have minus values.

> **Q23.** Similarly, what two variables are most positively correlated?

The most prominent positive correlation pairs i see are chocolate and winpercent, bar and chocolate, and nougat and bar. 

Principal Component Analysis
```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)

ggplotly(p)
```
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> **Q24.** What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard and pluribus. 

