---
title: "Class17_MiniProject"
author: "Angelita Rivera (PID A15522236)"
date: "11/23/2021"
output: 
  pdf_document:
    latex_engine: xelatex
---

# Mini-Project COVID Vaccination Rates

As we approach a period of travel and larger gatherings lets have a look at vaccination rates across the State. 

We will take data from the CA.gov site here: 
- “Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code"

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

## Ensure the date column is useful

We will use the **lubridate** package, which can make life allot easier when dealing with dates and times

> **Q1.** What column details the total number of people fully vaccinated? 

The column 'persons_fully_vaccinated'.

> **Q2.** What column details the Zip code tabulation area? 

The column 'zip_code_tabulation_area'.

> **Q3.** What is the earliest date in this dataset?

```{r}
vax$as_of_date[1]
```

> **Q4.** What is the latest date in this dataset?

```{r}
vax$as_of_date[nrow(vax)]
```

## Quick look at the data structure

As before we can use the **skim()** function to quickly overview and summarize the various columns of the dataset.
```{r}
skimr::skim(vax)
```

> **Q5.** How many numeric columns are in this dataset? 

There are nine numeric columns in this data set.

> **Q6.** Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

> **Q7.** What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

There is 10.00% of the persons_fully_vaccinate values missing. 

> **Q8.** [Optional]: Why might this data be missing?

They might be missing because of the military bases (or other areas) may not be contributing data. 


## Working with dates

```{r}
# install.packages("lubridate")

library(lubridate)
```


```{r}
today()
```

Here we make our data 'as_of_date' column lubridate format...

```{r}
# Specify that we are using the Year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

Now I can do useful math with dates more easily: 

> Q. How many days since the first entry?

```{r}
today() - vax$as_of_date[1]
```

> Q. How many days since the last entry?

```{r}
today() - vax$as_of_date[nrow(vax)]
```

> **Q9.** How many days between the first and last entry in the data set? 

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> **Q10.** How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```

This sounds good; 

```{r}
46*7
```


## Working with ZIP Codes

```{r}
#install.packages("zipcodeR")
#install.packages("terra")
```
```{r}
library(zipcodeR)

geocode_zip('92037')
```

More usefully, we can pull census data about ZIP code areas (including median household income etc.). For example:

```{r}
reverse_zipcode(c('92037', "92109") )
```

# Focus on San Diego County

Using base R;
```{r}
# Subset to San Diego county only areas
inds <- vax$county == "San Diego"

head(vax[inds,])
```

Using the **dplyr** package; 
```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

#How many entries are there in San Diego County?
nrow(sd)
```

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

> **Q11.** How many distinct zip codes are listed for San Diego County? 

```{r}
length(unique(sd$zip_code_tabulation_area))
```

> **Q12.** What San Diego County Zip code area has the largest 12 + Population in this dataset

```{r}
ind <- which.max(sd$age12_plus_population)
sd[ind, ]
```

> **Q.** What is the population in the 92037 ZIP Code area? 

```{r}
filter(sd, zip_code_tabulation_area == "92037")[1,]
```

> **Q13.** What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”? 

```{r}
sd.now <- filter(sd, as_of_date == "2021-11-09")

head(sd.now)
```

```{r}
mean(sd.now$percent_of_population_fully_vaccinated, na.rm = TRUE)
```

To get the 6-number summary; 

```{r}
summary(sd.now$percent_of_population_fully_vaccinated)
```

> **Q14.** Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

Using base R; 

```{r}
hist(sd.now$percent_of_population_fully_vaccinated)
```

Using ggplot; 

```{r}
library(ggplot2)

ggplot(sd.now) + 
  aes(percent_of_population_fully_vaccinated) + geom_histogram(bins = 15) + labs(title = "Histogram of Vaccination Rates Across San Diego County", subtitle = "As of 2021-11-09", x = "Percent of Population Fully Vaccinated in a Zip Code Area", y = "Count (Zip code areas")
```


# Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")

ucsd[1,]$age5_plus_population
```

> **Q15.** Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) +
  aes(as_of_date,
  percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(title = "Vaccination rate for La Jolla CA 92109", x = "Date", y="Percent Vaccinated")
```

We have about ~90% fully vaccinated. 

# Comparing 92037 to other similar sized areas

Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on as_of_date “2021-11-16”.

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36.all <- filter(vax, age5_plus_population > 36144)

nrow(vax.36.all)
```

How many unique zip codes have a population as large as 92037? 

```{r}
length(unique(vax.36.all$zip_code_tabulation_area))
```

> **Q16.** Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
#mean(vax.36$percent_of_population_fully_vaccinated)
```

Add H-line
```{r}
ggplot(ucsd) +
  aes(as_of_date,
  percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  geom_hline(yintercept = 0.66, col = "red") +
  ylim(c(0,1)) +
  labs(title = "Vaccination rate for La Jolla CA 92109", x = "Date", y="Percent Vaccinated")
```

> **Q17.** What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
summary(vax.36.all$percent_of_population_fully_vaccinated)
```

> **Q18.** Using ggplot generate a histogram of this data.

```{r}
hist(vax.36.all$percent_of_population_fully_vaccinated)
```

> **Q19.** Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

They are below the average value I calculated. 

> **Q20.** Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
ggplot(vax.36.all) + 
  aes(as_of_date, 
      percent_of_population_fully_vaccinated, 
      group = zip_code_tabulation_area) + 
      geom_line(alpha = 0.2, col = "blue") + 
      geom_hline(yintercept = 0.66, col = "red") + labs(title = "Vaccination rate across California", subtitle = "Only areas with a population above 36k are shown.", y = "Percent Vaccinated", x = "Date")
```

> **Q21.** How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?

I would rather not meet in person. I feel like we're going to see high rates of exposure, even if everyone is fully vaccinated. I would be happy to log online during next week's classes; or at least maybe we could have a hybrid week? Maybe Tuesday online and Thursday in person (after hopefully everyone gets tested); I do like the help in person provides, I'm just a little scared about potential exposure.












